### 2017-06-30 motivation.tex

Fig.~\ref{fig:new_order_code} shows a simplified version of the TPC-C
\C{new\_order} transaction.  TPC-C is a well-known Online Transaction
Processing (OLTP) benchmark that models an order-processing system for
a wholesale parts supply business. The business logic is captured in 5
database transactions that operate on 9 tables. \C{new\_order} is one
such transaction that operates on \C{District}, \C{Order},
\C{New\_order}, \C{Stock}, and \C{Order\_line} tables. The transaction
acts on the behalf of a customer, whose id is \C{c\_id}, to place a
new order for the given set of items (\C{item\_reqs}), to be served by
a warehouse under the district identified by \C{d\_id}. The
transaction does so by invoking appropriate SQL functionality,
captured by various calls to functions under the \C{SQL} module. All
\C{SQL} functions take the table name (a nullary constructor) as their
first argument. The higher-order \C{SQL.select1} function accepts a
boolean function that describes the selection criteria, and returns
any record that meets the criteria (it models the SQL query \C{SELECT
  \ldots\xspace LIMIT 1}). \C{SQL.update} also accepts a Boolean
function (3rd argument) to select the records to be updated. Its
2$%nd$ argument is a function that maps each selected record to a new
(updated) record. \C{SQL.insert} inserts a given record into the
specified table in the database.

@@

Because the aforementioned reasoning is reasonably simple to perform
manually, it may be feasible to apply techniques for concurrent
program verification~\cite{....} to automate much of this effort.
Serializability aids the tractability of verification by preventing
any interference among concurrently executing transactions.  Under
weak isolation\footnote{Weak isolation doesn't violate atomicity as
  long as the witnessed effects are those of committed transactions},
interferences of various kinds are permitted.  Although various forms
of weak isolation differ in the kinds of effects they allow and when
those effects are made visible, admitting interference of any kind
from concurrent atomic transactions makes verifying program invariants
significantly harder than under a serializable execution.

To illustrate, reconsider the behavior of the \C{new\_order}
transaction when executed with a \emph{Read Committed} (RC) isolation
level, the default isolation level in PostgreSQL, a widely used
open-source database system.  An RC transaction is isolated from
\emph{dirty writes}, i.e., writes of uncommitted transactions, but is
allowed to witness the writes of concurrent transactions as soon as
they are committed. Thus, with two concurrent instances of the
\C{new\_order} transaction (call them $T_1$ and $T_2$), both
concurrently placing new orders for different customers under the same
district (\C{d\_id}), PostgreSQL admits the two executions shown in
Fig.~\ref{fig:new_order_execs}.

@@

Notably, this example does not exhibit many common concurrency bugs
that arise from incorrect use of weak isolation such as
\emph{write-write} conflicts, or \emph{lost updates}.  While $T_1$ and
$T_2$ both increment the \C{d\_next\_o\_id} field of the district
record, they do so atomically (Line 4 of
Fig.~\ref{fig:new_order_code}), allowing both updates to be present in
the final state. Likewise, other well-known anomalies that
characterize RC isolation~\cite{berenson} such as \emph{fuzzy reads},
\emph{phantom reads}, \emph{read skew}, and \emph{write skew}, are
also not exhibited by the example.  Thus, program analyses that aim to
determine appropriate isolation by checking for possible
manifestations of these anomalies would fail to identify grounds for
promoting the isolation level of \C{new\_order} to something stronger.
Yet, if we take the semantics of the application into account, it is
quite clear that RC is not an appropriate isolation level for
\C{new\_order}. Clearly, the demonstrates the need for a high-level
verification framework that lets us reason about application semantics
in conjunction with the semantics of weak isolation.

While reasoning in terms of anomalies is cumbersome, and as the above
example shows often inadequate, reasoning about weak isolation in
terms of low-level operational details faces similar hurdles given the
conceptual gap between properties defined by consistency conditions
and data structures and state maintained by implementations.  To
overcome these challenges, we develop a framework that axiomatizes the
behavior of weak isolation implementations to enable modular reasoning
using well-established techniques such as Rely-Guarantee~\cite{...}
augmented to deal with weakly-isolated transactions.

To illustrate, consider how we might verify that \C{new_order} is
sound when executed under a \emph{Repeatable Read} (RR) isolation
level.  PostgreSQL executes an RR transaction by taking a
(conceptual) snapshot of the database state before the transaction
begins that is used by reads performing the transaction.  When the
transaction attempts to update a record, a check is made to determine
if the current version of the record in the database is same as the
snapshot version. If it is, an exclusive lock is obtained on the
record, and the update is performed. If the record is already locked,
the transaction blocks until the lock is released.  When the lock is
obtained, is held exclusively until the transaction commits or rolls
back. If the aforementioned version check fails (i.e., the database
has a later version than the snapshot), the transaction is rolled back
and re-executed.

Although this implementation, comprising many thousands of lines code,
is highly involved, its behavior in terms of state transitions on the
database is fairly simple.  Since transaction's reads are always
served from a snapshot, no state changes are witnessed while the
execution is in progress. Thus, insofar as an RR transaction is
concerned, the database state does not change during the execution.
Uncommitted writes are recorded in a transaction-local state.  When
the transaction commits, the local state is atomically written to the
global state to yield a global state that reflects the transaction's
updates.  However, unlike a strongly isolated serializable
transaction, the commit operation is performed against the current
state of the database, not the snapshot. Thus, after the transaction
finishes execution, but before it commits, the transaction is able to
witness the effects of all concurrent transactions.  The PostgreSQL RR
implementation effectively constrains this transition.

@@

The second observation that informs our approach is one that pertains
to automation.

\newpage

(\C{k-a1}), but before it commits, transaction \C{Wd2} (red) executes
and commits, writing the new balance (\C{k-a2}). {\sc rc} isolation
prevent \C{Wd2} from witnessing the uncommitted writes of transaction
\C{Wd1}.  Subsequently committing \C{Wd1} leads to the loss of
\C{Wd2}'s updates (the so-called \emph{lost update
anomaly}~\cite{berenson}), resulting in an incorrect balance of
\C{k-a1}. The execution on right describes a similar scenario with
\C{Wd1} and \C{Wd2} exchanging their roles.  Clearly, {\sc rc} is an
excessively weak isolation level for this program because it loses the
updates of one transaction, resulting in the violation of the
post-condition.  A stronger isolation level that prevents lost updates
is required. \iso{Snapshot Isolation} ({\sc si})~\cite{berenson} fits
this requirement; {\sc si} effectively serializes transactions that
update a shared data object by aborting and re-executing a transaction
if write-write conflicts are detected during its commit.\footnote{{\sc
si} however does not serialize transactions in the absence of
write-write conflicts.} Since {\sc si}, unlike {\sc ser}, does not
need expensive mechanisms such as lock-based concurrency control or
runtime monitoring, it is also more efficient, making it appropriate
for both withdraw transactions.

Thinking in terms of anomalies, as described above, is how database
programmers are often encouraged to reason about weak isolation.
Unfortunately, such reasoning does not rest on any sound foundation,
and is thus highly error-prone. Reasoning in terms of how weak
isolation variants are implemented is no better since it requires
programmers to understand low-level implementation details of the
database that are far removed from the application semantics. An
attractive  alternative in this context would be a principled
reasoning approach that combines declarative reasoning about isolation
guarantees with operational reasoning about programs. We demonstrate
how our proof system makes this possible in the context of the current
example.

First, we note that the example in Fig.~\ref{fig:motiv-eg-1} is a
concurrent program, and hence is amenable to \emph{rely-guarantee}
style reasoning~\cite{rgjones}, a compositional proof technique that
allows us to reason about the behaviour of individual threads by
abstracting away interferences induced by other threads (collectively
called \emph{the environment}) into a \emph{rely} relation. In an
ordinary concurrent program, every environment step is a valid
interference in the current thread. However, in the presence of
transactions executing under various levels of weak isolation,
determining what constitutes an interference is a non-trivial problem.
For example, a \iso{Serializable} transaction admits no interference,
whereas within a transaction executing under \iso{Read Uncommitted}
isolation, all interferences are valid. Between these two extremes are
various levels of isolation that admit some interferences while
prohibiting others. For example, \iso{Read Committed} isolation admits
interference of a committed transaction, but not those from
uncommitted transactions. \iso{Snapshot Isolation} admits interference
from committed non-conflicting transactions, and so on.

\begin{figure}
\begin{smathpar}
\begin{array}{lcl}
\psi_{RC} & \defeq & \forall T_1,T_2,\eta_1,\eta_2.\; \txn(\eta_1) = T_1 
  \conj \txn(\eta_2) = T_2 \\
  & & \hspace*{0.6in}\conj T_1 \neq T_2 \conj \eta_1 \hboar
  \eta_2 \Rightarrow T_1 \hboar \eta_2 \\
\psi_{SI} & \defeq & \forall T_1,T_2.\; T_1 \neq T_2 \conj
  (\exists \C{X}.~{T_1 \wrstoar \C{X}} \conj 
                      {T_2 \wrstoar \C{X}})\\
  &  & \hspace*{0.6in}\Rightarrow{T_1 \hboar T_2} \disj {T_2 \hboar T_1} \\
\end{array}
\end{smathpar}
\caption{\small Interference properties for different weak isolation
levels can be captured as constraints over the happens-before
relation.}
\label{fig:interference-ex}
\end{figure}

\begin{figure}
\centering
\subcaptionbox {
  \label{fig:hb-eg-1}
} [
  0.25\columnwidth
] {
  \includegraphics[scale=0.45]{Figures/hb-eg-1}
}
\subcaptionbox {
  \label{fig:hb-eg-2}
} [
  0.25\columnwidth
] {
  \includegraphics[scale=0.45]{Figures/hb-eg-2}
}
\subcaptionbox {
  \label{fig:hb-eg-3}
} [
  0.25\columnwidth
] {
  \includegraphics[scale=0.45]{Figures/hb-eg-3}
}
\subcaptionbox {
  \label{fig:hb-eg-4}
}{
  \includegraphics[scale=0.45]{Figures/hb-eg-4}
}

\caption{\small Possible executions for the example in
Fig.~\ref{fig:motiv-eg-1}. Letters R, W and C stand for read, write
and commit, respectively. Each blue dashed arrow represents a $\hbZ$
relationship between all operations of \C{Wd2} and an operation in
\C{Wd1}. $\psi_{SI}$ specification allows $\hbZ$ arrows (hence the
execution) in Fig.~\ref{fig:hb-eg-1}. $\psi_{RC}$ specification allows
$\hbZ$ arrows in
Figs.~\ref{fig:hb-eg-1},~\ref{fig:hb-eg-2},~\ref{fig:hb-eg-3}, and
~\ref{fig:hb-eg-4}.  }
\label{fig:motiv-eg-1-hb}
\vspace*{-8pt}
\end{figure}

For the rely-guarantee approach to be useful, it has to thus constrain
allowed interference in accordance with the chosen level of isolation
associated with a transaction. The first observation we make is that
we can constrain interference by constraining the nature of the
\emph{happens-before} ($\hbZ$) relation, which ultimately dictates
what other transactions become visible to a transaction, or its
constituents, and when. To do so, we axiomatize the $\hbZ$ relation to
capture the interference characteristics of different isolation levels
as their specifications. Two such specifications are
shown\footnote{Note that specifications presented here are solely for
illustration. The actual specifications described in
\S\ref{sec:ansi-isolation} are more nuanced.} in
Fig.~\ref{fig:interference-ex}. The \iso{Read Committed} specification
($\psi_{RC}$) allows an operation $\eta_1$ of a transaction $T_1$ to
happen-before an operation $\eta_2$ of another transaction $T_2$ only
if every operation in $T_1$ (including its commit) happens-before
$\eta_2$ (we let $T_1 \hboar \eta_2$ denote this). The specification
does not require $T_1$ to execute and commit before all operations of
$T_2$, thus allowing it to interfere in $T_2$, just as \C{Wd2}
interferes in \C{Wd1} in Fig.~\ref{fig:motiv-eg-1-a}. The
\iso{Snapshot Isolation} specification ($\psi_{SI}$) requires two
transactions that write to the same variable (denoted $\wrstoar$) to be
related by happens-before.  This effectively prohibits interference
due to actions in $T_1$ being interleaved in $T_2$ (or, vice versa).

Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ from \C{Wd2}
to \C{Wd1} allowed by $\psi_{RC}$ and $\psi_{SI}$. Arrows in all
executions are legal under $\psi_{RC}$ because in no execution does an
operation from \C{Wd2} happen-before an operation of \C{Wd1} without
the commit of \C{Wd2} also happening before the operation of \C{Wd1}.
$\psi_{SI}$ however disallows $\hbZ$ arrows (hence executions) in
Figs.~\ref{fig:hb-eg-2},~\ref{fig:hb-eg-3}, and~\ref{fig:hb-eg-4}
because such arrows establish $\hbZ$ edges between (operations in)
\C{Wd2} and only a subset of the operations in \C{Wd1}. Since the
result of an operation (\eg a read) depends on what happens before
that operation, the structure of the $\hbZ$ relation ultimately
dictates the final result of the program. The arrows in
Figs.~\ref{fig:hb-eg-1} and~\ref{fig:hb-eg-2} denote $\hbZ$
relationships that do not affect the value of \C{B} in a way that
causes the program to violate its post-condition. In contrast, arrows
in Figs.~\ref{fig:hb-eg-3} and~\ref{fig:hb-eg-4} denote $\hbZ$
relationships that lead to the violation of the post-condition. The
task of the reasoning framework is to determine if \emph{all} $\hbZ$
relationships allowed by an isolation level lead to the satisfaction
of the post-condition.

% There is a $\hb$ edge (first dashed) that is allowed by RC but
% prohibited by SI, which nonetheless does not lead to invariant
% violation.

The specification of an isolation level encodes its interference
characteristics as constraints over the $\hbZ$ relation. However, for
this to be useful in reasoning about programs, the rely-guarantee
framework should be able to use the specification to determine if an
interference is valid or not, allowing the programmer to only focus on
valid interferences. Our second observation is that this is possible
if the reasoning framework adequately tracks $\hbZ$ at each program
point, while preserving $\hbZ$ constraints as invariants between
program points. An interference that leads to the violation of $\hbZ$
constraints (i.e., an invalid interference) is thus automatically
prohibited. For instance, consider the program point after the write
to \C{B} in \C{Wd1}. The expected invariant ($\phi$) at that program
point is shown below ($\committed$ stands for ``committed''):

\begin{smathpar}
\begin{array}{c}
 \neg\committed(\C{Wd1}) \conj \C{Wd1} \wrstoar \C{B}  \conj 
    (\neg\committed(\C{Wd2}) \Rightarrow \C{B = k-a1}) \\
    \conj (\committed(\C{Wd2})
                \Rightarrow \C{B = k-a1-a2})
\end{array}
\end{smathpar}

\noindent $\phi$ asserts that \C{Wd1} is not yet committed, and that
it wrote to \C{B}, and the value of \C{B} is either \C{k-a1-a2} or
\C{k-a1} depending on whether or not \C{Wd2} is committed. If $\phi$
remains invariant until \C{Wd1} commits, then the post-condition (\C{B
= k-a1-a2}) can be established easily. However, an interference from
\C{Wd2} at this stage (captured by the last dashed arrow in
Fig.~\ref{fig:motiv-eg-1-hb}) may violate the invariant by writing
$\C{k-a2}$ to \C{B} and committing \C{Wd2}, thus leading to
$\committed(\C{Wd2}) \Rightarrow \C{B=k-a2}$. Fortunately,
\iso{Snapshot Isolation} prevents this interference, and this can be
shown by demonstrating that an interference from \C{Wd2} starting from
an execution state that satisfies $\psi_{SI} \wedge \phi$ leads to an
execution state where neither $\C{Wd2} \hboar \C{Wd1}$ nor $\C{Wd1}
\hboar \C{Wd2}$ holds; $\C{Wd2} \hboar \C{Wd1}$ does not hold because
\C{Wd1}'s write to \C{B} clearly happened before \C{Wd2}'s commit, and
$\C{Wd1} \hboar \C{Wd2}$ does not hold because \C{Wd1} has not yet
committed, while \C{Wd2} has already begun. Since \C{Wd2} also writes
to \C{B}, this violates the $\psi_{SI}$ constraint which we assume to
be an invariant. A proof that the post-condition holds now follows
from the contradiction. It is informative to note that if the
invariant is $\psi_{RC}$ instead of $\psi_{SI}$, we cannot derive a
contradiction and we cannot rule out the interference, which
(rightfully) causes  the proof to fail.

Implicit in the above discussion is the assumption of a strongly
consistent ({\sc sc}) store that guarantees the visibility of all
previously committed transactions. An {\sc sc} semantics can be built
into the reasoning framework, leading to a proof system tailor-made
for such stores.  However, for the reasoning framework to be truly
useful, it should be capable of handling different consistency
semantics, supporting stores that are weaker than {\sc sc} (e.g.,
causally or eventually consistent), and should be able to reconcile
conflicts between consistency and isolation constraints.  We
demonstrate how our reasoning framework makes this possible in the
following sections.

%% We have thus far assumed a strongly consistent ({\sc sc}) store that,
%% in the absence of transactions with special isolation requirements,
%% makes the effects of any operation immediately visible to all
%% subsequent operations. The natural behaviour of {\sc sc} to totally
%% order all operations w.r.t. the $\hbZ$ relation could be in conflict
%% with the constraints imposed by weak isolation. To meet the
%% requirements of weaker isolation levels, say {\sc rc}, the store has
%% to adapt itself to \emph{hide} the effects of concurrent transactions
%% until they are committed. Once committed though, effects need to be
%% immediately visible to subsequent operations. This semantics can be
%% built into the reasoning framework, leading to a proof system
%% tailor-made for such stores. However, for the reasoning framework to
%% be truly useful, it has to be parametric over different consistency
%% semantics, supporting stores that are weaker than {\sc sc} (e.g.,
%% causally or eventually consistent), and should be able to reconcile
%% conflicts between consistency and isolation constraints.  We
%% demonstrate how our reasoning framework makes this possible in the
%% following sections.

For example, \emph{Repeatable
  Read} is a stronger isolation level that RC; we might ask whether
executing the \C{new_order} transaction under this isolation level
would preserve the consistency conditions of the transaction.  To
answer this question, we must relate the low-level specification of RR
given in terms of e.g., trace-based
relations~\cite{adyaphd,gotsmanconcur15} to the semantics of TPC-C,
which is difficult.   Alternatively, one could reason about how RR is
implemented, say on PostgresSQL, and reason about the \C{new\_order}
implementation in conjunction with Postgres's RR
implementation. However, this is not any easier, considering that
Postgres employs a host of techniques, including Multiversion, locks
and rollbacks, to implement RR. The verification problem becomes even
more difficult if the aim is to facilitate automation. Our experience
suggests that a proof framework that requires programmers to relate
trace-based isolation specifications to program semantics necessarily
imposes heavy annotation burden, requires painstakingly long and
verbose proofs, and does not lend itself to automation. Weak isolation
aside, there are also the conventional challenges that face
(automatic) verification of concurrent programs: facilitating modular
proofs, inferring intermediary assertions and loop invariants that are
stable w.r.t interference, etc. A proof system for weakly isolated
transactions has to necessarily address these challenges, along with
those that arise due to weak isolation.

>>>>

Fig.~\ref{fig:new_order_code} shows a simplified version of the TPC-C
\C{new\_order} transaction.  TPC-C is a well-known Online Transaction
Processing (OLTP) benchmark that models an order-processing system for
a wholesale parts supply business. The business logic is captured in 5
database transactions that operate on 9 tables. \C{new\_order} is one
such transaction that operates on \C{District}, \C{Order},
\C{New\_order}, \C{Stock}, and \C{Order\_line} tables. The transaction
acts on the behalf of a customer, whose id is \C{c\_id}, to place a
new order for the given set of items (\C{item\_reqs}), to be served by
a warehouse under the district identified by \C{d\_id}. The
transaction does so by invoking appropriate SQL functionality,
captured by various calls to functions under the \C{SQL} module. All
\C{SQL} functions take the table name (a nullary constructor) as their
first argument. The higher-order \C{SQL.select1} function accepts a
boolean function that describes the selection criteria, and returns
any record that meets the criteria (it models the SQL query \C{SELECT
  \ldots\xspace LIMIT 1}). \C{SQL.update} also accepts a Boolean
function (3rd argument) to select the records to be updated. Its
2$^{nd}$ argument is a function that maps each selected record to a new
(updated) record. \C{SQL.insert} inserts a given record into the
specified table in the database.

@@

Because the aforementioned reasoning is reasonably simple to perform
manually, verfiying that soundess of TPC-C's consistency conditions
would appear to be feasible.  Serializability aids the tractability of
verification by preventing any interference among concurrently
executing transactions.  Under weak isolation\footnote{Weak isolation
  doesn't violate atomicity as long as the witnessed effects are those
  of committed transactions}, however, interferences of various kinds
are permitted.  Although the verification problem for weakly isolated
transactions would appear to be superficially similar to the
verification of (racy) concurrent programs~\cite{concurrentGC...},
weak isolation introduces new challenges arising from the use of
transactions, and new opportunities arising from the fact that the
store abstraction used by transaction is a relational database, not
low-level memory.  Our proof framework addresses these challenges and
exploits these opportunities to facilitiate fully automated
correctness proofs.

To illustrate, consider the behavior of the \C{new\_order} transaction
when executed with a \emph{Read Committed} (RC) isolation level, the
default isolation level in PostgreSQL, a widely used open-source
database system.  An RC transaction is isolated from \emph{dirty
  writes}, i.e., writes of uncommitted transactions, but is allowed to
witness the writes of concurrent transactions as soon as they are
committed. Thus, with two concurrent instances of the \C{new\_order}
transaction (call them $T_1$ and $T_2$), both concurrently placing new
orders for different customers under the same district (\C{d\_id}),
PostgreSQL admits the two executions shown in
Fig.~\ref{fig:new_order_execs}.

@@

Notably, this example does not exhibit any common concurrency bugs
that arise from incorrect use of weak isolation such as
\emph{write-write} conflicts, or \emph{lost updates}.  While $T_1$ and
$T_2$ both increment the \C{d\_next\_o\_id} field of the district
record, they do so atomically (Line 4 of
Fig.~\ref{fig:new_order_code}), allowing both updates to be present in
the final state. Likewise, other well-known anomalies that
characterize RC isolation~\cite{berenson} such as \emph{fuzzy reads},
\emph{phantom reads}, \emph{read skew}, and \emph{write skew}, are
also not exhibited by the example.  Thus, program analyses that aim to
determine appropriate isolation by checking for possible
manifestations of these anomalies would fail to identify grounds for
promoting the isolation level of \C{new\_order} to something stronger.
Yet, if we take the semantics of the application into account, it is
quite clear that RC is not an appropriate isolation level for
\C{new\_order}.

While reasoning in terms of anomalies is cumbersome, and as the above
example shows often inadequate, reasoning about weak isolation in
terms of low-level traces~\cite{adyaphd,gotsmanconcur15} confounds
high-level reasoning and automation.  Another alternative would
interleave weak isolation implementation details within the program,
yielding a (more-or-less) conventional concurrent program that can be
then subject to classical concurrent verification methods.
Considering the size and complexity of real-world transaction systems,
this approach is unlikely to scale.

In this paper, we adopt a different approach that \emph{lifts}
isolation semantics (\emph{not} their implementations) to the
application layer, providing a unified framework to simultaneously
reason about application invariants and isolation properties.  To
illustrate this idea informally, consider how we might verify that
$\C{new_order}$ is sound when executed under a \emph{Repeatable Read}
(RR) isolation level.  PostgreSQL executes an RR transaction by taking
a (conceptual) snapshot of the database state before the transaction
begins that is used by reads performing the transaction.  When the
transaction attempts to update a record, a check is made to determine
if the current version of the record in the database is same as the
snapshot version. If it is, an exclusive lock is obtained on the
record, and the update is performed. If the record is already locked,
the transaction blocks until the lock is released.  When the lock is
obtained, is held exclusively until the transaction commits or rolls
back. If the aforementioned version check fails (i.e., the database
has a later version than the snapshot), the transaction is rolled back
and re-executed.

Although this implementation, comprising many thousands of lines code,
is highly involved, its semantic behavior in terms of how it effects
transitions on the database state is fairly simple.  Since a
transaction's reads are always served from a snapshot, no state
changes are witnessed while the execution is in progress. Thus,
insofar as an RR transaction is concerned, the database state does not
change during the execution.  Uncommitted writes are recorded in a
transaction-local state.  When the transaction commits, the local
state is atomically written to the global state to yield a global
state that reflects the transaction's updates.  However, unlike a
strongly isolated serializable transaction, the commit operation is
performed against the current state of the database, not the
snapshot. Thus, after the transaction finishes execution, but before
it commits, the transaction is able to witness the effects of all
concurrent transactions.  The PostgreSQL RR implementation effectively
constrains this transition.

@@

The second observation that informs our approach is one that pertains
to automation. Program verification, even when machine-aided, often
entails significant annotation burden in the form of intermediary
assertions and loop invariants required to prove a program correct.
This is certainly true for concurrent program logics, such as
Rely-Gurantee, which extend Hoare logic with additional artifacts and
where (stable) intermediary assertions and loop invariants remain a
major source of annotation burden.  However, a relational database is
a significantly simpler abstraction than shared memory. There are no
pointers, or linked data structures, or aliasing.  Although a database
essentially abstracts a mutable state, the state is mutated through a
well-defined fixed number of interfaces (SQL statements), each tagged
with a logical formula describing what records are accessed and
updated.

This observation leads us away from thinking of database transactions
as concurrent imperative programs.  Instead, we see value in viewing
them as essentially functional computations that manage database state
monadically~\cite{statemonad}. We find it useful to reason about
statements that mutate the database state, not in terms of a pre- and
post-condition pair, but in terms of a state transformer that relates
the pre- and post-states of a statement. This state transformer
semantics can be defined algorithmically, just like predicate
transformer semantics (e.g., strongest post-condition).  Here, a state
transformer interprets a SQL statement in the set domain, taking
advantage of the fact that a a database is essentially a set of
tuples, and a SQL statement is a transformer over these sets.  The
benefit of this approach is that low-level loops can now be
substituted with higher-order combinators that automatically lift the
state transformer of its higher-order argument (i.e., the loop body)
to the state transformer of the combined expression (i.e., the loop).
Thus the semantics of a \C{foreach} loop, for instance, can be
captured as a state transformer, where the state is a set, and the
transformation defines a bind operation. We illustrate this intuition
on a simple example.

% \begin{figure}[!t]
% \centering
% %
% \begin{subfigure}[b]{0.46\textwidth}
% \begin{ocaml}
% Set s' = $\emptyset$;
% foreach x in s {
%   s'.add(f(x)); 
% }
% \end{ocaml}
% \caption{}
% \end{subfigure}
% %
% \begin{subfigure}[b]{0.5\textwidth}
% \begin{ocaml}
% let s' = ref [];
% foreach s (fun x -> s' := x::(!s'));
% \end{ocaml}
% \caption{Lorem ipsum, lorem ipsum,Lorem ipsum, lorem ipsum,Lorem ipsum}
% \end{subfigure}
% \caption{Caption place holder}
% %
% \caption{New versions are created from existing versions either
% through \C{push} or \C{merge}.}
% \label{fig:syntactic-ancestors}
% \end{figure}
% let s' = ref (Set.empty) in
% foreach xs @@ fun x -> 
%   begin
%     s' := Set.union !s' @@ 
%             Set.map_selected s (fun y -> y<x)
%                                (fun y -> y+x);
%     s' := Set.add !s' x;
%   end
\begin{figure}[!h]
\begin{ocaml}
foreach item_reqs @@ fun item_req -> do
  SQL.update Stock (fun s -> {s with s_qty = k1}) 
                   (fun s -> s.s_i_id = item_req.ol_i_id);
  SQL.insert Order_line {ol_o_id=k2; ol_d_id=k3; 
                         ol_i_id=item_req.ol_i_id; ol_qty=item_req.ol_qty}
\end{ocaml}
\caption{Foreach loop from Fig.~\ref{fig:new_order_code}}
\label{fig:foreach_code}
\end{figure}

Fig.~\ref{fig:foreach_code} shows a (simplified) snippet of code taken
from Fig.~\ref{fig:new_order_code}. Some irrelevant expressions have
been replaced with constants (\C{k1} to \C{k3}).  The body of the loop
executes a SQL update followed by an insert.  Recall that a transaction
reads from the global database ($\Delta$), and writes to a
transaction-local database ($\delta$) before committing these updates. An update
statement filters the tuples that match the search criteria from $\Delta$
and computes the updated tuples that are to be
added to the local database. Thus, its state transformer (call it
$T_U$) is the following function on sets\footnote{$\bind$ has higher precedence than $\cup$.}:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $\Delta$$\bind$(fun s -> if table(s) = stock && s.s_i_id = item_req.ol_i_id 
                                 then {{s with s_qty = k1}} (* a singleton set *)
                                 else $\emptyset$)
\end{ocaml}
% \begin{smathpar}
% \begin{array}{c}
%   \lambda\Delta.\lambda\delta.\, \delta \;\cup\; 
%       \Delta \,\bind\, (\lambda s.\, 
%           \C{if}\;{\C{stock}(s) \conj 
%                    \C{s.s\_i\_id} = \C{item\_req.ol\_i\_id}}\\
%            \hspace*{0.9in}
%            \C{then}\;{\{\{s \;\C{with}\; \C{s\_qty} = \C{k1}\}\}}\;
%            \C{else}\;{\emptyset})
% \end{array}
% \end{smathpar}
% \begin{verbatim}
%   Rmem(σ'(s')) = Rmem(σ(s')) U 
%                  Rmem[\x. Rmem[\y. if y<x then {y+x} else {}](σ(s)) U {x}](xs)
% \end{verbatim}
Observe that
$T_U(\Delta,\delta)$ is of the form $\delta \,\cup\,
F_U(\Delta)$. The transformer ($T_I(\Delta,\delta)$)
for the subsequent \C{insert} statement can be similarly calculated to
be of the form $\delta \cup F_I(\Delta)$.
Their composition gives the state transformer of the loop body:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ $F_U(\Delta) \,\cup\, F_I(\Delta)$
\end{ocaml}
Let $F_{body}(\Delta) = F_U(\Delta) \,\cup\, F_I(\Delta)$.   The
transformer for the \C{foreach} loop can now be computed as following:
\begin{ocaml}
  fun ($\Delta$,$\delta$) -> $\delta$ $\cup$ item_reqs$\bind$(fun item_req -> $F_{body}(\Delta)$)
\end{ocaml}
Observe that the transformer captures the precise semantics of the
loop as a formula in the domain of sets extended with a bind
function. The advantage in doing so is that we can now make use of a
semantics-preserving translation from this domain to first-order
logic, allowing us to leverage SMT solvers for automatic proofs
without having to infer loop invariant, or provide intermediate
assertions.  Sec.~\ref{sec:automation} describes this translation. In
the exposition thus far, we assumed $\Delta$ remains invaraint, which
is clearly not the case when we admit concurrency.  Necessary
concurrency extensions of the state transformer semantics is also
covered in Sec.~\ref{sec:automation}.  The immediate two sections,
however, focus on laying theoretical foundations for reasoning about
weak isolation.
